---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

```{r loading data, message=FALSE, warning=FALSE}

# Unzip and read dataset
unzip("activity.zip")
dataset = read.csv("activity.csv")

# Adjust date format
dataset$date = as.character(dataset$date)
dataset$date = as.Date(dataset$date, "%Y- %m- %d")

# Ignore NAs values
dataset_non_na = dataset[complete.cases(dataset),]

```

## What is mean total number of steps taken per day?

```{r steps per day, message=FALSE, warning=FALSE}

# Summarise steps per day
library(dplyr)
dataset_day = dataset_non_na %>%
        group_by(date) %>%
        summarize(steps_day = sum(steps, na.rm = TRUE))

# Plot histogram of steps per day
hist(dataset_day$steps_day,
     freq = T, 
     main = "Total number of steps taken each day",
     xlab = "Steps per day")

# Mean and Median of the total number os steps
mean_steps = as.character(round(mean(dataset_day$steps_day), 0))
median_steps = as.character(round(median(dataset_day$steps_day), 0)) 

```

The mean of steps taken each day: `r mean_steps`  
The median of steps taken each day: `r median_steps`


## What is the average daily activity pattern?

```{r steps per interval, message=FALSE, warning=FALSE}
# Summarise steps per 5-minute interval
dataset_interval = dataset_non_na %>%
        group_by(interval) %>%
        summarize(average_steps = mean(steps, na.rm = T))

# Plot time series of the 5-minute interval (x-axis) and the average number of steps taken
plot(dataset_interval$interval, dataset_interval$average_steps,
     type = "l",
     xlab = "5-minutes interval",
     ylab = "Average of steps taken",
     main = "Daily activity pattern",
     col =  "red",
     lwd = 1)

# Interval with max number of steps
max_interval = as.numeric(dataset_interval[dataset_interval$average_steps == max(dataset_interval$average_steps),1])

```

The 5-minute interval that contains the maximum number of steps is: `r max_interval`


## Imputing missing values

```{r missing values, message=FALSE, warning=FALSE}
# Calculate the total number of Missing values
missing_values = nrow(dataset[!complete.cases(dataset),])

# Replace Missing values values with mean value of the corresponding interval during all days
dataset_filled = dataset
dataset_filled = dataset_filled %>%
        group_by(interval) %>%
        mutate(steps = ifelse(is.na(steps), mean(steps, na.rm = T), steps))

# Group data by day
dataset_filled_day = dataset_filled %>%
        group_by(date) %>%
        summarize(steps_day = sum(steps, na.rm = TRUE))

# Plot a histogram of the total number of steps taken each day
hist(dataset_filled_day$steps_day,
     freq = T, 
     main = "Total number of steps taken each day",
     xlab = "Steps per day")     

# Mean and Median total number of steps taken per day
mean_steps_filled = as.character(round(mean(dataset_filled_day$steps_day), 0)) 
median_steps_filled = as.character(round(median(dataset_filled_day$steps_day), 0)) 

```

The total number of Missing values is `r missing_values` rows.  
The mean of steps taken each day with dataset filled missing values: `r mean_steps_filled`.  
The median of steps taken each day with dataset filled missing values: `r median_steps_filled`.  

The strategy to deal with missing values was replace them with the mean of specific interval. As we can see, the mean values remains equal in both methods, but the median value has slight increase. The total daily number os steps has a significant increase with this dataset filled.


## Are there differences in activity patterns between weekdays and weekends?

```{r plot weekdays, message=FALSE, warning=FALSE}
# Create factor variable weekdays
dataset_filled_weekdays = dataset_filled
dataset_filled_weekdays$days = weekdays(dataset_filled_weekdays$date)
dataset_filled_weekdays$fac_day = ifelse(dataset_filled_weekdays$days %in% c("sábado",
                                                                             "domingo"),
                                         "Weekend",
                                         "Weekday")
dataset_filled_weekdays$days = NULL

# Group data by interval
dataset_plot = dataset_filled_weekdays %>%
        group_by(interval, fac_day) %>%
        summarize(steps_day = mean(steps))

# Plot a time series plot of the 5-minute interval and the average number of steps taken
library(ggplot2)

h = ggplot(data = dataset_plot)
h + geom_line(aes(x = interval, y = steps_day, color = fac_day))+
  facet_wrap(~fac_day, ncol = 1)+
  labs(title = "Steps per interval Weekdays vs Weekends",
       y = "Steps per interval",
       x = "5-minutes interval",
       color = "Type of day")
```

It can be seen that on weekdays, average steps increase in the morning and in the end of day.
In weekends, the steps starting increase more lately, but the average values is more constant during the day. 






